Bayesian exponential random graph modeling of KWL Burial network

### 2020-12-04 analysis/paper/paper.Rmd
fixed typos

<!-- comparison between two networks using patchwork bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand if this network size different affects the posterior distributions, we conducted a sensitivity test by randomly removing `r diff_pre_post ` nodes from the post-European network to resemble the size of pre-European network with vertex bootstrap. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence interval for the network population, especially for smaller networks. We set the resampling size to 1,000 networks to compute network statistics, including density, clustering, and degree, for two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network density and degree, but higher extreme network clustering. This pattern is consistent before and after node removal for the post-European network, and the difference can be still identified between networks from different phases. This indicates our networks are relatively robust to differences in sample size and suggests a very small sample size effect.

Similarly, we can use vertex bootstrap to compare pre- and post-European networks since it allows a statistical testing for multiple networks of different sizes [@Chen2019; @Snijders1999]. We used t test for simulated networks and the results show that there are significant difference in all network statistics from both periods. In addition, since the prior setting will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

@@

<!-- theoretical framework for the results-->
The notable finding in our results is the change in network properties of Kiwulan burials from a more cohesive network (represented by higher transitivity and negative popularity) toward a more centralized network (lower transitivity and positive popularity) after the European presence during the 17th century. This confirms our hypothesized models of chronological differences in the structure of burials in terms of grave goods and burial attributes associated with demography and mortuary practices. The differences are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum, that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000]. Compared to vertical hierarchy that stresses stratified social characteristics, the corporate-network model is more effective to investigate the way how a small scale society organized and explore changes in social relation in a pericolonial context. A corporate-based society stresses shared power between individuals, integrative rituals, and social inequality, if any, would be associated with a social group or sector [@Siegel1999]. In contrast, a network-based society presents wealth accumulation through individual networks, resources control, and trade monopolization.

<!--archaeological implication based on the results-->
The covariants for both networks before and after the European presence provide further information of which variables relative to the formation of networks with the same prestige goods in common. We included age, sex, ritual pottery, and wealth index based on value of grave goods. The pre-European network indicates that burials with pottery as ritual vessel tend to be tied, while the burials within the same level of wealth tend to be connected in the post-European network. Moreover, the burials with ritual vessel tend not to be tied in the post-European network. This illustrates there are different mechanisms to form relations between burials in terms of the access to trade trade goods. Before the European arrival, ritual behavior seem to be an emphasized cultural practice relative to the use of prestige goods. After the European arrival, the cultural emphasis shifted from ritual to wealth difference. The demographic information, such as age and sex, do not show any correlations with the network formation for both periods. This may be influenced by some unidentified burials that limits a more insightful result. The distance between burials is not a variable that correlates to the network formation either.

@@

<!-- methodological implication-->
In this paper, we used burial network as a proxy to explore the social relations based on the argument that burials could be viewed as a social representation [@Binford1971; @Saxe1970]. We also note that the issue of disconnection between the role of the deceased in burial context and their real life, which can be resolved by examined different evidence out of the burial context [@Chapman2003]. The ERGM network analysis within a Bayesian framework provides a robust way to examine relations between burials based on grave goods and attributes associated with demography and mortuary practices. The adoption of Bayesian inference on ERGM can solve the issues of computational tractability and degeneracy in parameter estimation in ERGM [@Caimo2017]. In addition, a Bayesian framework reduced the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. We demonstrated a plausible and robust method to test the hypothesized model of chronological changes in burial structures with probabilistic interpretation. Although some network statistics of covariants do not showing the pattern as we assumed, the Bayesian approach is an efficient and effective method for evaluating the endogenous network effects relative to the overall structure of burials. This sturcutre can be viewed as a representation of a given society.

>>>>

<!-- comparison between two networks using patchwork bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand if this network size difference affects the posterior distributions, we conducted a sensitivity test by randomly removing `r diff_pre_post ` nodes from the post-European network to resemble the size of pre-European network with vertex bootstrap. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence interval for the network population, especially for smaller networks. We set the resampling size to 1,000 networks to compute network statistics, including density, clustering, and degree, for two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network density and degree, but higher extreme network clustering. This pattern is consistent before and after node removal for the post-European network, and the difference can be still identified between networks from different phases. This indicates our networks are relatively robust to differences in sample size and suggests a very small sample size effect.

Similarly, we can use vertex bootstrap to compare pre- and post-European networks since it allows statistical testing for multiple networks of different sizes [@Chen2019; @Snijders1999]. We used t-test for simulated networks and the results show that there are significant differences in all network statistics from both periods. In addition, since the prior setting will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

@@

<!-- theoretical framework for the results-->
The notable finding in our results is the change in network properties of Kiwulan burials from a more cohesive network (represented by higher transitivity and negative popularity) toward a more centralized network (lower transitivity and positive popularity) after the European presence during the 17th century. This confirms our hypothesized models of chronological differences in the structure of burials in terms of grave goods and burial attributes associated with demography and mortuary practices. The differences are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum, that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000]. Compared to vertical hierarchy that stresses stratified social characteristics, the corporate-network model is more effective to investigate the way how a small scale society organized and explore changes in social relations in a pericolonial context. A corporate-based society stresses shared power between individuals, integrative rituals, and social inequality, if any, would be associated with a social group or sector [@Siegel1999]. In contrast, a network-based society presents wealth accumulation through individual networks, resource control, and trade monopolization.

<!--archaeological implication based on the results-->
The covariants for both networks before and after the European presence provide further information of which variables relative to the formation of networks with the same prestige goods in common. We included age, sex, ritual pottery, and wealth index based on the value of grave goods. The pre-European network indicates that burials with pottery as ritual vessels tend to be tied, while the burials within the same level of wealth tend to be connected in the post-European network. Moreover, the burials with ritual vessels tend not to be tied in the post-European network. This illustrates there are different mechanisms to form relations between burials in terms of access to trade goods. Before the European arrival, ritual behavior seem to be an emphasized cultural practice relative to the use of prestige goods. After the European arrival, the cultural emphasis shifted from ritual to wealth difference. The demographic information, such as age and sex, do not show any correlations with the network formation for both periods. This may be influenced by some unidentified burials that limits a more insightful result. The distance between burials is not a variable that correlates to the network formation either.

@@

<!-- methodological implication-->
In this paper, we used burial data as a proxy to explore the social relations based on the argument that burials could be viewed as a social representation [@Binford1971; @Saxe1970]. We also note that the issue of disconnection between the role of the deceased in the burial context and their real life, which can be resolved by examined different evidence out of the burial context [@Chapman2003]. The ERGM network analysis within a Bayesian framework provides a robust way to examine relations between burials based on grave goods and attributes associated with demography and mortuary practices. The adoption of Bayesian inference on ERGM can solve the issues of computational tractability and degeneracy in parameter estimation in ERGM [@Caimo2017]. In addition, a Bayesian framework reduced the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. We demonstrated a plausible and robust method to test the hypothesized model of chronological changes in burial structures with probabilistic interpretation. Although some network statistics of covariants do not show the pattern as we assumed, the Bayesian approach is an efficient and effective method for evaluating the endogenous network effects relative to the overall structure of burials. This structure can be viewed as a representation of a given society.
